Calibrating MEM seeding heuristics

### 2018-05-13 MEM.tex
Swapped v and w. Removed head row from M.
Fixed errors in the theory. Cleaned symbols to math code.

The weighted generating function of the transition from $(\uparrow)_j$ to
$\Downarrow$ is
\begin{equation}
D_{\gamma-j}(z) = \omega z \sum_{i=0}^{\gamma-j-1} (qz)^i.
\end{equation}

The weighted generating function of the transition from
$(\uparrow)_j$ to $\downarrow$ is
\begin{equation}
\tilde{D}_{\gamma-j}(z) = \tilde{\omega} z
  \sum_{i=0}^{\gamma-j-1} (qz)^i.
\end{equation}

@@

If the next terminator is $\Downarrow$, there is a limit on the
number of $\square$ symbols. Indeed, if there are $\gamma$ or more
nucleotides between the two $\Downarrow$ symbols, they constitute a shared
MEM seed. As mentioned in section~\ref{sec:trans_from_up}, the weighted
generating function of the $\Downarrow$ symbol is $pz(1-\mu/3)^N$. By the
same rationale as above, the weighted generating function of the
transition from $\Downarrow$ to $\Downarrow$ is
\begin{equation}
A_\gamma(z) = \omega z
  \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
\end{equation}

If the next terminator is $\downarrow$ we have to dinstinguish two cases.
If there are $\gamma-1$ or fewer error-free nucleotides, we must only make
sure that there is no $\uparrow$ symbol among them. The weighted
generating function is
\begin{equation}
\tilde{\omega} z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
\end{equation}

There can also be more than $\gamma-1$ error-free nucleotides, but in this
case, we must make sure that at least one potentially masking thread masks
the main thread.

@@

Finally, the weighted generating function of the transition from
$\Downarrow$ to $\downarrow$ is
\begin{equation}
\tilde{A}_\infty(z) =
\tilde{\omega} z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i +
pz\sum_{i=\gamma}^\infty (1 - \alpha_i^N) \cdot (qz)^i.
\end{equation}

@@

If there are $m = 0$ potentially masking threads, the next terminator
cannot be $\sim$. Assuming that there are $m > 0$ potentially masking
threads at the $\downarrow$ symbol, the probability that the have not all
faied at position $i$ is $\xi_{i,m}$. Similarly, the probability that the
last strictly masking thread fails at position $i$ is
$\xi_{i-1,N-m}-\xi_{i,N-m}$. So the probability that the next terminator
is $(\sim)_i$ is $q^j V_i$, where
\begin{eqnarray}
\label{eq:Si}
V_i &=& \sum_{m=0}^{N-1} Q_m \cdot \xi_{i,m} \cdot
  (\xi_{i-1,N-m}-\xi_{i,N-m}) \notag \\
  &=& \frac{\alpha_i^N-\alpha_{i-1}^N-\gamma_i^N+\delta_{i-1}^N}
{1-(1-\mu/3)^N} \text{, where}
\end{eqnarray}

@@

We have assumed $m>0$, but the sum above can start at $m=0$ because
$\xi_{i,0}=0$. The weighted generating function of the transition from
$\downarrow$ to $(\sim)_i$ is
\begin{equation}
v_i(z) = \frac{\alpha_i^N-\alpha_{i-1}^N-
  \gamma_i^N+\delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
\end{equation}

@@

Referring again to the decision of Fig.~\ref{fig:decision_tree}, we see
that the $\uparrow$ symbol occurs in an error-free stretch, if all the
potentially masking threads fail before or at the same nucleotide as the
strictly masking threads. Assuming that there are $m \geq 0$ potentially
masking threads at the $\downarrow$ symbol, the probability that they have
all failed at position $i$ is $1-\xi_{i,m}$. As we have seen above, the
probability that the last strictly masking thread fails at position $j$ is
$\xi_{i-1,N-m}-\xi_{i,N-m}$. So the probability that the next terminator
is the $\uparrow$ symbol at position $i$ is $q^i W_i$, where
\begin{eqnarray}
  W_i &=& \sum_{m=0}^{N-1} Q_m \cdot (1-\xi_{i,m}) \cdot
  (\xi_{i-1,N-m}-\xi_{i,N-m}) \notag \\
    &=& \frac{\gamma_i^N-\delta_{i-1}^N}{1-(1-\mu/3)^N}.
\end{eqnarray}

The weighted generating function of the transition from $\downarrow$ to
$(\uparrow)_i$ is
\begin{equation}
w_i(z) = \frac{\gamma_i^N- \delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
\end{equation}

%In order to compute the weighted generating functions of the other
%transitions, first observe that
%\begin{equation}
%V_j+W_j = \frac{\alpha_j^N - \alpha_{j-1}^N}{1-(1-\mu/3)^N}.
%\end{equation}
%So the probability that an error-free stretch of length $i \geq 1$
%following a $\downarrow$ symbol contains neither a $\sim$ symbol nor a
%$\uparrow$ symbol is
%\begin{equation}
%1 - \sum_{j=1}^i V_j+W_j = 1 - \frac{\alpha_i^N -
%\alpha_0^N}{1-(1-\mu/3)^N}.
%\end{equation}
%
%The probability that the next terminator after a $\downarrow$ symbol is a
%$\Downarrow$ symbol at position $j$ is the probability of an error-free
%stretch of size $j-1$ that contains neither the $\sim$ symbol nor the
%$\uparrow$ symbol -- probability as above multiplied by $q^j$ -- followed
%by a $\Downarrow$ symbol -- probability $pz(1-\mu/3)^N$. Thus, the
%weighted generating function of the transition from $\downarrow$ to
%$\Downarrow$ is
%\begin{equation}
%B_\infty(z) = \omega z \sum_{i=0}^\infty
%\Big(1 - \frac{\alpha_i^N - \alpha_0^N}{1-(1-\mu/3)^N} \Big) (qz)^i.
%\end{equation}

@@

So the weighted generating function of the transition from $\downarrow$ to
$\Downarrow$ is
\begin{equation}
B_\infty(z) = \omega z\sum_{i=0}^\infty
\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
\end{equation}

By the same rationale, we see that the weighted generating function of the
transition from $\downarrow$ to $\downarrow$ is
\begin{equation}
\tilde{B}_\infty(z) = \tilde{\omega} z\sum_{i=0}^\infty
\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
\end{equation}

@@

The sum of the terms above for $m = 1, \ldots, N-1$ is $V_j$ from equation
(\ref{eq:Si}), which is the probability of the $(\sim)_j$ terminator.

@@

The probability that the last of the $m$ potentially masking threads fails
at position $i$ after entering state $(\sim)_j$ is
$\xi_{i-1,m}-\xi_{i,m}$. Dividing by $V_j$ and summing over $m = 1,
\ldots, N-1$, we find that the weighted generating function of the
transition from $(\sim)_j$ to $(\uparrow)_{j+i}$ is
\begin{eqnarray}
y_{j,i}(z) &=& \sum_{m=1}^{N-1} \frac{P_{m,j}}{V_j}
\big( \xi_{i-1,m}-\xi_{i,m} \big) (qz)^i \notag \\
&=& \frac{\beta_{j,i+j}^N - \beta_{j,i+j-1}^N
-\beta_{j-1,i+j}^N + \beta_{j-1,i+j-1}^N}
  {\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i,
\end{eqnarray}
where $\alpha_j$, $\delta_j$, $\gamma_j$ are as above, and where
\begin{equation}
\beta_{j,i} = 1 - (1-\mu)^j\mu/3 - (1-\mu)^i(1-\mu/3).
\end{equation}

The probability that the next terminator is a $\Downarrow$ symbol or a
$\downarrow$ symbol at position $i$ is the probability that at least one
of the potentially masking threads survives until position $i-1$ --
probability $\xi_{i-1,m}$ -- and that the next nucleotide is a sequencing
error -- probability $p$. Dividing by $V_j$ and summing over $m = 1,
\ldots, N-1$, we find that the weighted generating function of the
transtion from $(\sim)_j$ to
$\Downarrow$ is
\begin{eqnarray}
C_{\gamma-j}(z) &=& \omega z \sum_{i=1}^{\gamma-j-1} \sum_{m=1}^{N-1}
  \frac{P_{m,j}}{V_j} \xi_{i-1,m} \cdot (qz)^{i-1} \notag \\
  &=& \omega z \sum_{i=1}^{\gamma-j-1}
\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j-1}^N
  +\beta_{j-1,i+j-1}^N }
{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^{i-1}.
\end{eqnarray}

With the same notations, the weighted generating function from $(\sim)_j$
to $\downarrow$ is seen to be 
\begin{equation}
\tilde{C}_{\gamma-j}(z) = \tilde{\omega} z \sum_{i=1}^{\gamma-j-1}
\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j-1}^N
  +\beta_{j-1,i+j-1}^N }
{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^{i-1}.
\end{equation}

@@

For the $\downarrow$ symbol, the weighted generating function is
\begin{equation}
T_{\downarrow}(z) = \sum_{i=0}^\infty
\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
\end{equation}

For the $(\uparrow)_j$ state, the weighted generating function is
\begin{equation}
T_{(\uparrow)_j}(z) = \sum_{i=0}^{\gamma-j-1} (qz)^i.
\end{equation}

For the other symbols, we must take care that the potentially masking
threads do not all fail (otherwise this would imply the presence of a
$\uparrow$ terminator in the tail). For the $\Downarrow$ symbol, the
weighted generating function is
\begin{equation}
T_{\Downarrow}(z) = \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
\end{equation}

For the $(\sim)_j$ state, the weighted generating function is
\begin{equation}
T_{(\sim)_j}(z) = \sum_{i=0}^{\gamma-j-1}
\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j}^N +\beta_{j-1,i+j}^N }
{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i.
\end{equation}

In summary,
\begin{equation*}
T(z) = 
\begin{blockarray}{cc}
\begin{block}{c[c]}
\DN & T_\Downarrow(z) \\
\dn & T_\downarrow(z) \\
\eq{1} & T_{(\sim)_1}(z) \\
\vdots & \vdots \\
\eq{\gamma-1} & T_{(\sim)_{\gamma-1}}(z) \\
\up{1} & T_{(\uparrow)_1}(z) \\
\vdots & \vdots \\
\up{\gamma-1} & \; T_{(\uparrow)_{\gamma-1}}(z) \; \\
\end{block}
\end{blockarray}.
\end{equation*}

@@

\begin{equation*}
\begin{blockarray}{ccccccccc}
   & \DN & \dn &\eq{1} & \ldots & \eq{\gamma-1} &
    \up{1} & \ldots & \up{\gamma-1} \\
\begin{block}{c[cccccccc]}
\DN & A_\gamma(z) & \tilde{A}_\infty(z) & 0 & \ldots & 0 & u_1(z)
    & \ldots & u_{\gamma-1}(z) \\
\dn & B_\infty(z) & \tilde{B}_\infty(z) & v_1(z) & \ldots &
    v_{\gamma-1}(z) & w_1(z) & \ldots & w_{\gamma-1}(z) \\
\eq{1} & C_{\gamma-1}(z) & \tilde{C}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
    \ldots & y_{1,\gamma-2}(z) \\
\eq{2} & C_{\gamma-2}(z) & \tilde{C}_{\gamma-2}(z) & 0 & \ldots & 0 & 0 &
    \ldots & y_{2,\gamma-3}(z) \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
    \ddots & \vdots \\
\eq{\gamma-1} & C_1(z) & \tilde{C}_1(z) & 0 & 
    \ldots & 0 & 0 & \ldots & 0 \\
\up{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0 \\
\up{2} & D_{\gamma-2}(z) & \tilde{D}_{\gamma-2}(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0  \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
    \ddots & \vdots \\
\up{\gamma-1} & D_1(z) & \tilde{D}_1(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0 \\
\end{block}
\end{blockarray}
\end{equation*}

In the case $N = 1$, reads never contain the symbol $\sim$ and the matrix
is
\begin{equation*}
\begin{blockarray}{cccccc}
   & \DN & \dn & \up{1} & \ldots & \up{\gamma-1} \\
\begin{block}{c[ccccc]}
\DN & A_\gamma(z) & \tilde{A}_\infty(z) & u_1(z)
    & \ldots & u_{\gamma-1}(z) \\
\dn & B_\infty(z) & \tilde{B}_\infty(z) & w_1(z)
    & \ldots & w_{\gamma-1}(z) \\
\up{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) & 0 &
    \ldots & 0 \\
\up{2} & D_{\gamma-2}(z) & \tilde{D}_{\gamma-2}(z) & 0 &
    \ldots & 0  \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
\up{\gamma-1} & D_1(z) & \tilde{D}_1(z) & 0 & \ldots & 0 \\
\end{block}
\end{blockarray}
\end{equation*}

>>>>

The weighted generating function of the transition from $(\uparrow)_j$ to
$\Downarrow$ is
\begin{equation}
C^\Downarrow_j(z) = \omega z \sum_{i=0}^{\gamma-j-1} (qz)^i.
\end{equation}

The weighted generating function of the transition from
$(\uparrow)_j$ to $\downarrow$ is
\begin{equation}
C^\downarrow_j(z) = \tilde{\omega} z
  \sum_{i=0}^{\gamma-j-1} (qz)^i.
\end{equation}

@@

If the next terminator is $\Downarrow$, there is a limit on the
number of $\square$ symbols. Indeed, if there are $\gamma$ or more
nucleotides between the two $\Downarrow$ symbols, they constitute a shared
MEM seed. As mentioned in section~\ref{sec:trans_from_up}, the weighted
generating function of the $\Downarrow$ symbol is $pz(1-\mu/3)^N$. By the
same rationale as above, the weighted generating function of the
transition from $\Downarrow$ to $\Downarrow$ is
\begin{equation}
A^\Downarrow(z) = \omega z
  \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
\end{equation}

If the next terminator is $\downarrow$ we have to dinstinguish two cases.
In the example below, there is a shared MEM seed only if $\gamma \leq
11$. The main thread is unmasked, so the number of nucleotides between the
error symbols matters.
\begin{inset}
\includegraphics[scale=.9]{ddown_to_down_case_1.pdf}
\end{inset}
In this other example, there can be no MEM seed because the first
duplicate masks the main thread. This depends neither on the value of
$\gamma$ nor on the number of nucleotides between the error symbols.
\begin{inset}
\includegraphics[scale=.9]{ddown_to_down_case_2.pdf}
\end{inset}

So if the $\Downarrow$ and the $\downarrow$ symbols are separated by
$\gamma-1$ or fewer nucleotides, there is no MEM seed and we must only
rule out the presence of other terminators in the segment (in this case
the $\uparrow$ symbol). The weighted generating function is
\begin{equation}
\tilde{\omega} z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
\end{equation}

On the other hand, if the $\Downarrow$ and the $\downarrow$ symbols are
separated by $\gamma$ or more nucleotides, there could be a shared MEM
seed, so we must enforce the condition that at least one potentially
masking thread masks the main thread up to and including the $\downarrow$
symbol (this incidentally rules out the presence of other terminators in
the segment).

@@

Finally, the weighted generating function of the transition from
$\Downarrow$ to $\downarrow$ is
\begin{equation}
A^\downarrow(z) =
\tilde{\omega} z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i +
pz\sum_{i=\gamma}^\infty (1 - \alpha_i^N) \cdot (qz)^i.
\end{equation}

@@

If there are $m = 0$ potentially masking threads, the next terminator
cannot be $\sim$. Assuming that there are $m > 0$ potentially masking
threads at the $\downarrow$ symbol, the probability that the have not all
faied at position $i$ is $\xi_{i,m}$. Similarly, the probability that the
last strictly masking thread fails at position $i$ is
$\xi_{i-1,N-m}-\xi_{i,N-m}$. So the probability that the next terminator
is $(\sim)_i$ is $q^j W_i$, where
\begin{eqnarray}
\label{eq:Si}
W_i &=& \sum_{m=0}^{N-1} Q_m \cdot \xi_{i,m} \cdot
  (\xi_{i-1,N-m}-\xi_{i,N-m}) \notag \\
  &=& \frac{\alpha_i^N-\alpha_{i-1}^N-\gamma_i^N+\delta_{i-1}^N}
{1-(1-\mu/3)^N} \text{, where}
\end{eqnarray}

@@

We have assumed $m>0$, but the sum above can start at $m=0$ because
$\xi_{i,0}=0$. The weighted generating function of the transition from
$\downarrow$ to $(\sim)_i$ is
\begin{equation}
w_i(z) = \frac{\alpha_i^N-\alpha_{i-1}^N-
  \gamma_i^N+\delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
\end{equation}

@@

Referring again to the decision of Fig.~\ref{fig:decision_tree}, we see
that the $\uparrow$ symbol occurs in an error-free stretch, if all the
potentially masking threads fail before or at the same nucleotide as the
strictly masking threads. Assuming that there are $m \geq 0$ potentially
masking threads at the $\downarrow$ symbol, the probability that they have
all failed at position $i$ is $1-\xi_{i,m}$. As we have seen above, the
probability that the last strictly masking thread fails at position $j$ is
$\xi_{i-1,N-m}-\xi_{i,N-m}$. So the probability that the next terminator
is the $\uparrow$ symbol at position $i$ is $q^i W_i$, where
\begin{eqnarray}
  V_i &=& \sum_{m=0}^{N-1} Q_m \cdot (1-\xi_{i,m}) \cdot
  (\xi_{i-1,N-m}-\xi_{i,N-m}) \notag \\
    &=& \frac{\gamma_i^N-\delta_{i-1}^N}{1-(1-\mu/3)^N}.
\end{eqnarray}

The weighted generating function of the transition from $\downarrow$ to
$(\uparrow)_i$ is
\begin{equation}
v_i(z) = \frac{\gamma_i^N- \delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
\end{equation}

@@

So the weighted generating function of the transition from $\downarrow$ to
$\Downarrow$ is
\begin{equation}
B^\Downarrow(z) = \omega z\sum_{i=0}^\infty
\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
\end{equation}

If the next terminator is $\downarrow$, we again have to distinguish two
cases. In the example below there is no MEM seed, but the nature of the
terminator depends on $\gamma$. If $\gamma < 10$, the state $(\sim)_9$
does not exist, so there is a single segment of size 12, terminated by
$\downarrow$. If $\gamma \geq 10$, the state $(\sim)_9$ exists and there
is a segment of size 9 terminated by $\sim$, followed by a segment of size
3 terminated by $\downarrow$.
\begin{inset}
\includegraphics[scale=.9]{down_to_down_case_1.pdf}
\end{inset}

So if the two $\downarrow$ symbols are separated by $\gamma-1$ or fewer
nucleotides, there is no MEM seed and we must only rule out the presence
of other terminators in the segment (in this case $\sim$ and $\uparrow$).
With the probability computed above that there are neither $\sim$ nor
$\uparrow$ symbols, the weighted generating function is
\begin{equation}
\tilde{\omega} z\sum_{i=0}^{\gamma-1}
\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
\end{equation}

If the two $\downarrow$ symbols are separated by $\gamma$ or more
nucleotides, there could be a shared MEM seed, so we must enforce the
condition that the main thread is masked. In addition, we must also rule
out the presence of the $\sim$ terminator among the first $\gamma-1$
nucleotides (but not among the remaining nucleotides). Finally, we must
rule out the presence of the $\uparrow$ terminator throughout the segment.
This is satisfied if and only if $i.$ a strictly masking thread survives
until the penultimate nucleotide of the fragment or $ii.$ all the strictly
masking threads fail before the penultimate nucleotide but not before the
$\gamma$-th nucleotide, and a potentially masking thread survives until
the last nucleotide of the fragment.

If there are $m$ potentially masking threads and $N-m$ strictly masking
threads, the respective probabilities of conditions $i.$ and $ii.$ are
$\xi_{i-1,N-m} \cdot (1-(1-\mu/3)^N)$ and $(\xi_{i-1,N-m}
-\xi_{\gamma-1,N-m})(1-\alpha_{i-1}^m)$. Since the conditions are
are mutually exclusive, the full probability is
\begin{equation}
\sum_{m=0}^{N-1}Q_m \bigg(\xi_{i-1,N-m} \cdot (1-(1-\mu/3)^N) +
(\xi_{\gamma-1,N-m}-\xi_{i-1,N-m})(1-\alpha_{i-1}^m) \bigg).
\end{equation}

The first term is equal to $1-\alpha_{i-1}^N$, the second is equal to
\begin{equation}
\frac{\alpha_{i-1}^N - \alpha_{\gamma-1}^N -
  \zeta_{i-1,i-1}^N + \zeta_{\gamma-1,i-1}^N}{1-(1-\mu/3)^N}.
\end{equation}
where
\begin{equation}
\zeta_{j,i} = 1-(1-\mu)^j\mu/3-(1-\mu)^i\mu/3(1-\mu/3)
\end{equation}

Finally, the weighted generating function of the transitions from
$\downarrow$ to $\downarrow$ is

\begin{equation*}
\begin{split}
B^\downarrow(z) = \tilde{\omega} z \sum_{i=0}^{\gamma-1}
& \frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i + \\
pz\sum_{i=\gamma}^\infty & \left( 1-\alpha_i^N + 
\frac{\alpha_i^N - \alpha_{\gamma-1}^N -
  \zeta_{i,i}^N + \zeta_{\gamma-1,i}^N}{1-(1-\mu/3)^N} \right)
(qz)^i
\end{split}
\end{equation*}

@@

The sum of the terms above for $m = 1, \ldots, N-1$ is $W_j$ from equation
(\ref{eq:Si}), which is the probability of the $(\sim)_j$ terminator.

@@

The probability that the last of the $m$ potentially masking threads fails
at position $i$ after entering state $(\sim)_j$ is
$\xi_{i-1,m}-\xi_{i,m}$. Dividing by $W_j$ and summing over $m = 1,
\ldots, N-1$, we find that the weighted generating function of the
transition from $(\sim)_j$ to $(\uparrow)_{j+i}$ is
\begin{eqnarray}
y_{j,i}(z) &=& \sum_{m=1}^{N-1} \frac{P_{m,j}}{W_j}
\big( \xi_{i-1,m}-\xi_{i,m} \big) (qz)^i \notag \\
&=& \frac{\beta_{j,i+j}^N - \beta_{j,i+j-1}^N
-\beta_{j-1,i+j}^N + \beta_{j-1,i+j-1}^N}
  {\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i,
\end{eqnarray}
where $\alpha_j$, $\delta_j$, $\gamma_j$ are as above, and where
\begin{equation}
\beta_{j,i} = 1 - (1-\mu)^j\mu/3 - (1-\mu)^i(1-\mu/3).
\end{equation}

The probability that the next terminator is a $\Downarrow$ symbol or a
$\downarrow$ symbol at position $i$ is the probability that at least one
of the potentially masking threads survives until position $i-1$ --
probability $\xi_{i-1,m}$ -- and that the next nucleotide is a sequencing
error -- probability $p$. Dividing by $W_j$ and summing over $m = 1,
\ldots, N-1$, we find that the weighted generating function of the
transtion from $(\sim)_j$ to
$\Downarrow$ is
\begin{eqnarray}
D^\Downarrow_j(z) &=& \omega z \sum_{i=0}^{\gamma-j-2} \sum_{m=1}^{N-1}
  \frac{P_{m,j}}{W_j} \xi_{i,m} \cdot (qz)^i \notag \\
  &=& \omega z \sum_{i=0}^{\gamma-j-2}
\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j}^N
  +\beta_{j-1,i+j}^N }
{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i.
\end{eqnarray}

With the same notations, the weighted generating function from $(\sim)_j$
to $\downarrow$ is seen to be 
\begin{eqnarray}
D^\downarrow_j(z) = \tilde{\omega} z
  \sum_{i=0}^{\gamma-j-2} \sum_{m=1}^{N-1}
  \frac{P_{m,j}}{W_j} \xi_{i,m} \cdot (qz)^i +
  pz \sum_{i=\gamma-j-1}^\infty \sum_{m=1}^{N-1}
  \frac{P_{m,j}}{W_j} (1-\alpha_i^m) \cdot (qz)^i \notag \\
\tilde{\omega} z \sum_{i=0}^{\gamma-j-2}
\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j}^N
  +\beta_{j-1,i+j}^N }
{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i + \notag \\
pz \sum_{i=\gamma-j-1}^\infty
\frac{ \alpha_j^N-\alpha_{j-1}^N-\zeta_{j,i+j}^N
  +\zeta_{j-1,i+j}^N }
{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i + \notag \\
\end{eqnarray}

@@

For the $\downarrow$ symbol, the weighted generating function is
\begin{equation}
T^\downarrow(z) = \sum_{i=0}^\infty
\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
\end{equation}

For the $(\uparrow)_j$ state, the weighted generating function is
\begin{equation}
T^\uparrow_j(z) = \sum_{i=0}^{\gamma-j-1} (qz)^i.
\end{equation}

For the other symbols, we must take care that the potentially masking
threads do not all fail (otherwise this would imply the presence of a
$\uparrow$ terminator in the tail). For the $\Downarrow$ symbol, the
weighted generating function is
\begin{equation}
T^\Downarrow(z) = \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
\end{equation}

For the $(\sim)_j$ state, the weighted generating function is
\begin{equation}
T^\sim_j(z) = \sum_{i=0}^{\gamma-j-1}
\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j}^N +\beta_{j-1,i+j}^N }
{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i.
\end{equation}

In summary,
\begin{equation*}
T(z) = 
\begin{blockarray}{cc}
\begin{block}{c[c]}
\DN & T^\Downarrow(z) \\
\dn & T^\downarrow(z) \\
\up{\gamma-1} & T^\uparrow_{\gamma-1}(z) \\
\vdots & \vdots \\
\up{1} & \; T^\uparrow_1(z) \; \\
\eq{\gamma-1} & T^\sim_{\gamma-1}(z) \\
\vdots & \vdots \\
\eq{1} & T^\sim_1(z) \\
\end{block}
\end{blockarray}.
\end{equation*}

@@

\begin{equation*}
\begin{blockarray}{ccccccccc}
   & \DN & \dn &\up{\gamma-1} & \ldots & \up{1} &
    \eq{\gamma-1} & \ldots & \eq{1} \\
\begin{block}{c[cccccccc]}
\DN & A_\gamma(z) & \tilde{A}_\infty(z) & u_{\gamma-1}(z) & \ldots &
    u_1(z) & 0 & \ldots & 0 \\
\dn & B_\infty(z) & \tilde{B}_\infty(z) & v_{\gamma-1}(z) & \ldots &
    v_1(z) & w_{\gamma-1}(z) & \ldots & w_1(z) \\
\up{\gamma-1} & C_1(z) & \tilde{C}_1(z) & 0 & 
    \ldots & 0 & 0 & \ldots & 0 \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
    \ddots & \vdots \\
\up{1} & C_{\gamma-1}(z) & \tilde{C}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0 \\
\eq{\gamma-1} & D_1(z) & \tilde{D}_1(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0 \\
\eq{\gamma-2} & D_2(z) & \tilde{D}_2(z) & y_{\gamma-2,1}(z) &
    \ldots & 0 & 0 & \ldots & 0 \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
    \ddots & \vdots \\
\eq{2} & D_{\gamma-2}(z) & \tilde{D}_{\gamma-2}(z) &
    y_{2,\gamma-3}(z) & \ldots & 0 & 0 & \ldots & 0 \\
\eq{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) &
    y_{1,\gamma-2}(z) & \ldots & 0 & 0 & \ldots & 0 \\
\end{block}
\end{blockarray}
\end{equation*}

In the case $N = 1$, reads never contain the symbol $\sim$ and the matrix
is
\begin{equation*}
\begin{blockarray}{cccccc}
   & \DN & \dn & \up{1} & \ldots & \up{\gamma-1} \\
\begin{block}{c[ccccc]}
\DN & A_\gamma(z) & \tilde{A}_\infty(z) & u_1(z)
    & \ldots & u_{\gamma-1}(z) \\
\dn & B_\infty(z) & \tilde{B}_\infty(z) & v_1(z)
    & \ldots & v_{\gamma-1}(z) \\
\up{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) & 0 &
    \ldots & 0 \\
\up{2} & D_{\gamma-2}(z) & \tilde{D}_{\gamma-2}(z) & 0 &
    \ldots & 0  \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
\up{\gamma-1} & D_1(z) & \tilde{D}_1(z) & 0 & \ldots & 0 \\
\end{block}
\end{blockarray}
\end{equation*}
